---
title: "2026 ASA Actuarial Case Competition"
output: html_document
date: "2026-09-14"
---

# Setting Up

```{r}
# load packages
library(tidyverse)
library(statmod)
library(tweedie)

# load dataset
data <- read_csv("06 - CAS Predictive Modeling Case Competition- Dataset.csv")
head(data)
```


```{r}
glimpse(data)
```

# EDA

## Among Tiers

Source (for def of Preferred, Standard and Non-Standard Tiers):
https://www.investopedia.com/terms/p/privatepassenger-auto-insurance-policyholder-risk-profile.asp

Preferred: least risky; least likely to file a claim
Standard: average risk
Non-Standard: riskiest; most likely to file a claim

Judging by average claim amount alone, with the riskiest tier having the highest average claim amount, we determine that:
  
  - Tier 1: Preferred
  - Tier 2: Standard
  - Tier 3: Non-Standard

```{r}
data |> 
  group_by(risk_tier) |>
  summarize(num_policyholders = length(unique(name)),
            count = length(name),
            num_nonzero_claims = sum(amount > 0),
            prop_nonzero_claims = round(num_nonzero_claims/count, 2),
            num_sprinklered = sum(sprinklered),
            prop_sprinklered = round(num_sprinklered/count, 2),
            avg_amount = round(mean(amount), 2),
            avg_dist = round(mean(distance_to_campus), 3)) |>
  pivot_longer(cols = -risk_tier, names_to = "metric", values_to = "value") |> 
  pivot_wider(names_from = risk_tier, values_from = value)
```

```{r}
data |> 
  group_by(risk_tier) |>
  summarize(avg_claim_amt = mean(amount))
```

```{r}
data |> 
  filter(amount > 0) |>
  group_by(risk_tier) |>
  summarize(avg_nonzero_claim_amt = mean(amount))
```

Tier 2's average claim amount is roughly the same as the average claim amount of the entire dataset. On the other hand, Tier 1 has below average claim amount, and Tier 3 has above average claim amount. 

```{r}
mean(data$amount)
mean(data[data$amount > 0, ]$amount)
```

### Average Claim Amount per Policyholder within Subgroups

```{r}
# Personal Property Protection
data |> 
  filter(coverage == "Personal Property") |>
  summarize(avg_amount = mean(amount), .by = c(risk_tier, student_id)) |>
  group_by(risk_tier) |>
  summarize(avg_amount_per_policyholder = mean(avg_amount))
```

## Data Validation

No claim amounts seem to exceed the insurance's maximum coverage. 

```{r}
# checking if every claim amt aligns with insurance coverage max

# Personal Property Protection (up to $10,000)
data |> 
  filter(coverage == "Personal Property") |>
  pull(amount) |>
  max()

# Liability Protection (up to $500,000)
data |> 
  filter(coverage == "Liability") |>
  pull(amount) |>
  max()

# Guest Medical Protection (up to $150,000)
data |> 
  filter(coverage == "Guest Medical") |>
  pull(amount) |>
  max()
```

# Model Training

## Data Preprocessing

The given data set has 40,071 rows, but we are told that there are only 10,000 policyholders and 4 different coverage types. Since there's no duplicated rows, this means there are some policyholders who filed more than one claim. Thus, for policyholders who filed more than one claim per insurance coverage, we need to combine the amount for those claims. 

```{r}
# check if there's any duplicate rows
cat("There are ", sum(duplicated(data)), "duplicated rows.")
```

## GLM to Predict Expected Claim Amount (Full Model)

Source: https://metricgate.com/blogs/tweedie-glm-insurance-pure-premium/

### Personal Property

```{r}
training <- data |> filter(training_holdout == "training", 
                           coverage == "Personal Property")
holdout <- data |> filter(training_holdout == "holdout",
                          coverage == "Personal Property")
```

```{r}
# estimate Tweedie power index p

profile <- tweedie_profile(
  amount ~ class + study + gpa + greek + off_campus +
    distance_to_campus + gender + sprinklered,
  data = training, p.vec = seq(1.1, 1.6, by = 0.1),
  link.power = 0, do.plot = FALSE
)

p_hat <- profile$p.max
cat("Estimated Tweedie power p:", round(p_hat, 3))

# fits Tweedie GLM w/ log link
fit <- glm(
  amount ~ class + study + gpa + greek + off_campus + 
    distance_to_campus + sprinklered,
  data = training,
  family = tweedie(var.power = p_hat, link.power = 0),
  control = glm.control(maxit = 15)
)
```

```{r}
fit
```

```{r}
holdout$pred  <- predict(fit, newdata = holdout, type = "response")
training$pred <- fitted(fit)

avg_amount <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
avg_amount
```

```{r}
head(holdout[, c("student_id", "coverage", "risk_tier", "amount", "pred")], 50)
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
avg_amount |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Personal Property Claim Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

## Variable Selection

Since there are more samples than variables ($n = 40071 > p = 11$), we use Backward Stepwise Selection to perform variable selection. 

```{r}

```

# Algorithm for Pricing

```{r}

```












