---
title: "2026 ASA Actuarial Case Competition"
output: html_document
date: "2026-09-14"
---

# Setting Up

```{r}
# load packages
library(tidyverse)
library(statmod)
library(tweedie)

# load dataset
data <- read_csv("06 - CAS Predictive Modeling Case Competition- Dataset.csv")
head(data)
```


```{r}
glimpse(data)
```

# Model Training

## Data Preprocessing

The given data set has 40,071 rows, but we are told that there are only 10,000 policyholders and 4 different coverage types. Since there's no duplicated rows, this means there are some policyholders who filed more than one claim. Thus, for policyholders who filed more than one claim per insurance coverage, we need to combine the amount for those claims. 

```{r}
# check if there's any duplicate rows
cat("There are ", sum(duplicated(data)), "duplicated rows.")
```

```{r}
#Combining Multiple Claims made by the same policyholder and covergae
data <- data |>
  group_by(student_id,coverage) |>
  summarize(
    amount= sum(amount, na.rm = TRUE),
    name= first(name),
    class=first(class),
    study=first(study),
    gpa=first(gpa),
    greek= first(greek),
    off_campus= first(off_campus),
    distance_to_campus = first(distance_to_campus),
    gender= first(gender),
    sprinklered =first(sprinklered),
    risk_tier= first(risk_tier),
    training_holdout= first(training_holdout),
    .groups = "drop"
)
```

```{r}

# Grouping the continous variables into 3 bands to make the pricing model easier to rate 
data <- data |>
  mutate(
    gpa_interval= case_when(
      gpa<2 ~ "Sub 2.0",
      gpa<3 ~ "2.0 to 2.99",
      gpa>=3 ~"Greater than 3"
    ),

  distance_interval= case_when(
    distance_to_campus==0 ~ "Living On Campus",
    distance_to_campus <=1 ~ "Living 1 mile or less from Campus",
    distance_to_campus > 1 ~ " Living more than 1 mile from Campus"
  )
)

```
    

## GLM to Predict Expected Claim Amount (Full Model)

Source: https://metricgate.com/blogs/tweedie-glm-insurance-pure-premium/

### Personal Property

```{r}
training <- data |> filter(training_holdout == "training", 
                           coverage == "Personal Property")
holdout <- data |> filter(training_holdout == "holdout",
                          coverage == "Personal Property")
```

```{r}
# estimate Tweedie power index p

profile <- tweedie_profile(
  amount ~ class + study + gpa + greek + off_campus +
    distance_to_campus + gender + sprinklered,
  data = training, p.vec = seq(1.1, 1.6, by = 0.1),
  link.power = 0, do.plot = FALSE
)

p_hat <- profile$p.max
cat("Estimated Tweedie power p:", round(p_hat, 3))

# fits Tweedie GLM w/ log link
fit <- glm(
  amount ~ class + study + gpa + greek + off_campus + 
    distance_to_campus + sprinklered,
  data = training,
  family = tweedie(var.power = p_hat, link.power = 0),
  control = glm.control(maxit = 15)
)
```

```{r}
fit
```

```{r}
holdout$pred  <- predict(fit, newdata = holdout, type = "response")
training$pred <- fitted(fit)

avg_amount <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
avg_amount
```

```{r}
head(holdout[, c("student_id", "coverage", "risk_tier", "amount", "pred")], 50)
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
avg_amount |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Personal Property Claim Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```
## Guest Medical
```{r}
# Seperating the Guest Medical Data into the training and holdout sets
training <-data |> filter(training_holdout == "training",coverage== "Guest Medical")

holdout<- data |> filter(training_holdout == "holdout", coverage == "Guest Medical")
```

```{r}
# Estinmating the Tweedie Power Index 
profile <- tweedie_profile(
  amount ~ class + study + gpa_interval + greek + off_campus + distance_interval + gender + sprinklered,
  data= training, p.vec=seq(1.1,1.6, by= 0.1),
  link.power=0, do.plot= FALSE
)

p_est<- profile$p.max
cat("Tweedie Power P Estimation:", round(p_est,3))

# Using Log Link to Fit the Tweeide GLM  
fit <- glm(
  amount ~ class + study + gpa_interval + greek + off_campus+ distance_interval + gender + sprinklered,
  data = training,
  family= tweedie(var.power = p_est, link.power=0),
  control = glm.control(maxit= 15)
)
```
```{r}
fit
```

```{r}
holdout$pred <- predict(fit, newdata=holdout, type ="response")
training$pred <- fitted(fit)

avg_amount <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), fitted= mean(pred), difference= abs(observed -fitted))
avg_amount
```
```{r}
head(holdout[,c("student_id","coverage","risk_tier","amount","pred")],50)
```
```{r}
#Creating a Bar Graph to Compare the Average Claim amount against the predicted claim amount

avg_amount |>
  pivot_longer(cols= c(observed,fitted), names_to="type",values_to="mean_pp") |>
  mutate(type = factor(type, levels= c("observed","fitted"), labels= c("Observed","Fitted"))) |>
  ggplot(aes(x= risk_tier, y= mean_pp, fill= type)) + geom_col(position= "dodge") +
  scale_fill_manual(values = c("Observed"="Blue","Fitted"="Red")) +
  labs(title= "Observed Results Vs Tweedie-fitted \n Average Guest Medical Claim Amount",
  x="Risk Tier", y= "Mean Pure Premium", fill ="")
```
### Liability 
```{r}
training <- data |>
  filter(training_holdout=="training", coverage=="Liability")
holdout <- data |> 
filter(training_holdout== "holdout", coverage == "Liability")
```

```{r}
#Estimating the Tweedie Power Index
profile <- tweedie_profile(
  amount ~ class + study+ gpa_interval+greek+off_campus+distance_interval+ gender+sprinklered,
  data=training,
  p.vec=seq(1.1,1.6,by=0.1),
  link.power =0,
  do.plot= FALSE
)

p_est <- profile$p.max
cat("Estimated Tweedie Power P:", round(p_est,3))
```
```{r}
# Using a Log link to fit the Tweedie GLM
fit<- glm(
  amount ~ class+ study+ gpa_interval+ greek+ off_campus+ distance_interval+ gender+sprinklered,
  data=training,
  family=tweedie(
    var.power=p_est,
    link.power=0
),
control=glm.control(maxit = 15)
)
```

```{r}
fit

```

```{r}
# Predicting the expected claim amounts for holdout observations
holdout$pred <- predict(
  fit,
  newdata= holdout,
  type= "response"
)

training$pred <- fitted(fit)
```
```{r}

avg_amount <- holdout |>
  group_by(risk_tier) |>
  summarize(
    observed = mean(amount),
    fitted = mean(pred),
    difference= abs(observed-fitted)
  )

avg_amount
```

```{r}
head(
  holdout[, c("student_id", "coverage", "risk_tier", "amount", "pred")],
  50
)
```

```{r}

avg_amount |>
  pivot_longer(
    cols=c(observed,fitted),
    names_to="type",
    values_to= "mean_pp"
  ) |>

  mutate(
    type=factor(
      type,
      levels= c("observed","fitted"),
      labels = c("Observed", "Fitted")
  )
) |>

  ggplot(aes(
    x= risk_tier,
    y= mean_pp,
    fill= type
  )) +
  geom_col(position="dodge")+
  scale_fill_manual(
    values= c(
      "Observed"= "Blue",
      "Fitted"= "Red"
    )
  ) +
  labs(
    title= "Observed Results vs. Tweedie-Fitted \n Average Liability Claim Amount",
    x= "Risk Tier",
    y= "Mean Pure Premium",
    fill= ""
  )
  
```


## Variable Selection

Since there are more samples than variables ($n = 40071 > p = 11$), we use Backward Stepwise Selection to perform variable selection. 

```{r}

```

# Algorithm for Pricing

```{r}

```












