---
title: "2026 ASA Actuarial Case Competition"
output: html_document
date: "2026-09-14"
---

# Setting Up

```{r}
# load packages
library(tidyverse)

# load data set
data <- read_csv("06 - CAS Predictive Modeling Case Competition- Dataset.csv")
head(data)
```


```{r}
glimpse(data)
```

# EDA

## Among Tiers

Source (for def of Preferred, Standard and Non-Standard Tiers):
https://www.investopedia.com/terms/p/privatepassenger-auto-insurance-policyholder-risk-profile.asp

Preferred: least risky; least likely to file a claim
Standard: average risk
Non-Standard: riskiest; most likely to file a claim

Judging by average claim amount alone, with the riskiest tier having the highest average claim amount, we determine that:
  
  - Tier 1: Preferred
  - Tier 2: Standard
  - Tier 3: Non-Standard

```{r}
data |> 
  group_by(risk_tier) |>
  summarize(num_policyholders = length(unique(name)),
            count = length(name),
            num_nonzero_claims = sum(amount > 0),
            prop_nonzero_claims = round(num_nonzero_claims/count, 2),
            num_sprinklered = sum(sprinklered),
            prop_sprinklered = round(num_sprinklered/count, 2),
            avg_amount = round(mean(amount), 2),
            avg_dist = round(mean(distance_to_campus), 3)) |>
  pivot_longer(cols = -risk_tier, names_to = "metric", values_to = "value") |> 
  pivot_wider(names_from = risk_tier, values_from = value)
```

```{r}
data |> 
  group_by(risk_tier) |>
  summarize(avg_claim_amt = mean(amount))
```

```{r}
data |> 
  filter(amount > 0) |>
  group_by(risk_tier) |>
  summarize(avg_nonzero_claim_amt = mean(amount))
```

Tier 2's average claim amount is roughly the same as the average claim amount of the entire dataset. On the other hand, Tier 1 has below average claim amount, and Tier 3 has above average claim amount. 

```{r}
mean(data$amount)
mean(data[data$amount > 0, ]$amount)
```

### Average Claim Amount per Policyholder within Subgroups

```{r}
# Personal Property Protection
data |> 
  filter(coverage == "Personal Property") |>
  summarize(avg_amount = mean(amount), .by = c(risk_tier, student_id)) |>
  group_by(risk_tier) |>
  summarize(avg_amount_per_policyholder = mean(avg_amount))
```

## Claim Amounts

The amount column is heavily right-skewed (zero-inflated). Thus, it's not normally distributed. 

```{r}
hist(data$amount, freq = FALSE, breaks = 100, 
     main = "Distribution of Claim Amount", xlab = "Amount")
lines(density(data$amount), col = "red", lwd = 2)
```

The nonzero claim amounts are heavily right-skewed.  

```{r}
amount <- data |> 
  filter(amount > 0) |> 
  pull(amount)

hist(amount,
     main = "Distribution of Nonzero Claim Amount")
```

By performing log transformation, the nonzero claim amounts distribution appear to be skew normal. 

```{r}
hist(log(amount),
     main = "Distribution of Log Transformed Nonzero Claim Amount")
```

## Data Validation

No claim amounts seem to exceed the insurance's maximum coverage. 

```{r}
# checking if every claim amt aligns with insurance coverage max

# Personal Property Protection (up to $10,000)
data |> 
  filter(coverage == "Personal Property") |>
  pull(amount) |>
  max()

# Liability Protection (up to $500,000)
data |> 
  filter(coverage == "Liability") |>
  pull(amount) |>
  max()

# Guest Medical Protection (up to $150,000)
data |> 
  filter(coverage == "Guest Medical") |>
  pull(amount) |>
  max()
```

# Model Training

## GLM to Predict Expected Claim Amount

```{r}
colnames(data)
```

```{r}
head(data)
```

```{r}
# training_data <- data |> filter(training_holdout == "training")
# glm(amount ~ class + study + gpa + greek, data = training_data, 
#     family = Gamma(link = "log"))
```

# Algorithm for Pricing

```{r}

```












