---
title: "EDA"
output: html_document
date: "2026-09-28"
---

# Setting Up

```{r}
# load packages
library(tidyverse)
library(statmod)
library(tweedie)

# load dataset
data <- read_csv("06 - CAS Predictive Modeling Case Competition- Dataset.csv")
head(data)
```

# Data Preprocessing

```{r}
#Combining Multiple Claims made by the same policyholder and coverage
data <- data |>
  group_by(student_id,coverage) |>
  summarize(
    amount= sum(amount, na.rm = TRUE),
    name= first(name),
    class=first(class),
    study=first(study),
    gpa=first(gpa),
    greek= first(greek),
    off_campus= first(off_campus),
    distance_to_campus = first(distance_to_campus),
    gender= first(gender),
    sprinklered =first(sprinklered),
    risk_tier= first(risk_tier),
    training_holdout= first(training_holdout),
    .groups = "drop"
)

head(data)
```

# EDA

## Among Tiers

Source (for def of Preferred, Standard and Non-Standard Tiers):
https://www.investopedia.com/terms/p/privatepassenger-auto-insurance-policyholder-risk-profile.asp

Preferred: least risky; least likely to file a claim
Standard: average risk
Non-Standard: riskiest; most likely to file a claim

Judging by average claim amount alone, with the riskiest tier having the highest average claim amount, we determine that:
  
  - Tier 1: Preferred
  - Tier 2: Standard
  - Tier 3: Non-Standard

```{r}
data |> 
  group_by(risk_tier) |>
  summarize(num_policyholders = length(unique(name)),
            count = length(name),
            num_nonzero_claims = sum(amount > 0),
            prop_nonzero_claims = round(num_nonzero_claims/count, 2),
            num_sprinklered = sum(sprinklered),
            prop_sprinklered = round(num_sprinklered/count, 2),
            avg_amount = round(mean(amount), 2),
            avg_dist = round(mean(distance_to_campus), 3)) |>
  pivot_longer(cols = -risk_tier, names_to = "metric", values_to = "value") |> 
  pivot_wider(names_from = risk_tier, values_from = value)
```

```{r}
data |> 
  group_by(risk_tier) |>
  summarize(avg_claim_amt = mean(amount))
```

```{r}
data |> 
  filter(amount > 0) |>
  group_by(risk_tier) |>
  summarize(avg_nonzero_claim_amt = mean(amount))
```

Tier 2's average claim amount is roughly the same as the average claim amount of the entire dataset. On the other hand, Tier 1 has below average claim amount, and Tier 3 has above average claim amount. 

```{r}
mean(data$amount)
mean(data[data$amount > 0, ]$amount)
```

### Average Claim Amount per Policyholder within Subgroups

```{r}
# Personal Property Protection
data |> 
  filter(coverage == "Personal Property") |>
  summarize(avg_amount = mean(amount), .by = c(risk_tier, student_id)) |>
  group_by(risk_tier) |>
  summarize(avg_amount_per_policyholder = mean(avg_amount))
```

# Data Validation

No claim amounts seem to exceed the insurance's maximum coverage. 

```{r}
# checking if every claim amt aligns with insurance coverage max

# Personal Property Protection (up to $10,000)
data |> 
  filter(coverage == "Personal Property") |>
  pull(amount) |>
  max()

# Liability Protection (up to $500,000)
data |> 
  filter(coverage == "Liability") |>
  pull(amount) |>
  max()

# Guest Medical Protection (up to $150,000)
data |> 
  filter(coverage == "Guest Medical") |>
  pull(amount) |>
  max()
```

# Data Visualizations


## Distribution of Claim Amounts

```{r}
hist(data$amount,
     breaks = 30,
     col = "thistle",
     xlab = "Amount",
     main = "Distribution of All Claim Amounts")
```

```{r}
hist(data |>
       filter(amount > 0) |>
       pull(amount),
     breaks = 50,
     col = "palegreen3",
     xlab = "Claim Amount",
     main = "Distribution of Nonzero Claim Amounts")
```

```{r}
hist(log(data |> 
           filter(amount > 0) |>
           pull(amount)),
     breaks = 20,
     col = "aquamarine3",
     xlab = "log(Claim)",
     main = "Distribution of Log-Transformed Claim Amounts")
```



# ALE segmenting
```{r}
data |>
  filter(coverage == "Additional Living Expense") |>
  mutate(gpa_group = cut(
    gpa,
    breaks = seq(0, 4, by = 0.5),
    include.lowest = TRUE
  )) |>
  group_by(student_id, risk_tier, gpa_group) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(risk_tier, gpa_group) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = round(mean(had_claim), 3),
    avg_amount_per_policyholder = round(mean(total_amount), 2),
    avg_loss_per_claimant = round(mean(total_amount[had_claim]), 2),
    .groups = "drop"
  )
```
```{r}


gpa_summary <- data |>
  filter(
    coverage == "Additional Living Expense",
    training_holdout == "training"
  ) |>
  mutate(gpa_group = cut(
    gpa,
    breaks = c(0, 1, 3, 4),
    include.lowest = TRUE
  )) |>
  group_by(student_id, risk_tier, gpa_group) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(risk_tier, gpa_group) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )
gpa_summary

```
```{r}
ggplot(
  gpa_summary,
  aes(
    x = gpa_group,
    y = avg_amount_per_policyholder
  )
) +
  geom_col() +
  facet_wrap(~ risk_tier, ncol = 1) +
  labs(
    title = "ALE Average Loss per Policyholder by GPA and Risk Tier",
    x = "GPA Range",
    y = "Average ALE Loss per Policyholder"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```
```{r}
gpa_all_coverages <- data |>
  filter(training_holdout == "training") |>
  mutate(gpa_group = cut(
    gpa,
    breaks = c(0, 1, 3, 4),
    include.lowest = TRUE
  )) |>
  group_by(student_id, coverage, risk_tier, gpa_group) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, gpa_group) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

```


```{r}

ggplot(
  gpa_all_coverages,
  aes(
    x = gpa_group,
    y = avg_amount_per_policyholder
  )
) +
  geom_col() +
  facet_grid(coverage ~ risk_tier, scales = "free_y") +
  labs(
    title = "Average Loss per Policyholder by GPA, Coverage, and Risk Tier",
    x = "GPA Range",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal()
```
```{r}
gpa_all_coverages |>
  select(
    coverage,
    risk_tier,
    gpa_group,
    num_policyholders,
    num_with_claim
  )

```

# OTHER VARIABLES
# DISTANCE TO CAMPUS

```{r}
data |>
  filter(training_holdout == "training") |>
  summarize(
    min_distance = min(distance_to_campus),
    q1 = quantile(distance_to_campus, 0.25),
    median = median(distance_to_campus),
    q3 = quantile(distance_to_campus, 0.75),
    max_distance = max(distance_to_campus),
    mean_distance = mean(distance_to_campus)
  )

data |>
  filter(training_holdout == "training") |>
  distinct(student_id, .keep_all = TRUE) |>
  ggplot(aes(x = distance_to_campus)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Distance to Campus",
    x = "Distance to Campus",
    y = "Number of Policyholders"
  ) +
  theme_minimal()

data |>
  filter(
    training_holdout == "training",
    distance_to_campus > 0
  ) |>
  distinct(student_id, .keep_all = TRUE) |>
  summarize(
    num_students = n(),
    min = min(distance_to_campus),
    q1 = quantile(distance_to_campus, 0.25),
    median = median(distance_to_campus),
    q3 = quantile(distance_to_campus, 0.75),
    max = max(distance_to_campus)
  )
```


#testing ALE for distance_to_campus

```{r}
distance_summary <- data |>
  filter(
    coverage == "Additional Living Expense",
    training_holdout == "training"
  ) |>
  mutate(distance_group = cut(
    distance_to_campus,
    breaks = c(-Inf, 0, 1, 3, 6, Inf),
    labels = c("0", "0-1", "1-3", "3-6", "6+")
  )) |>
  group_by(student_id, risk_tier, distance_group) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(risk_tier, distance_group) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

ggplot(
  distance_summary,
  aes(
    x = distance_group,
    y = avg_amount_per_policyholder
  )
) +
  geom_col() +
  facet_wrap(~ risk_tier, ncol = 1) +
  labs(
    title = "ALE Average Loss per Policyholder by Distance to Campus",
    x = "Distance to Campus",
    y = "Average ALE Loss per Policyholder"
  ) +
  theme_minimal()

distance_summary


```
# trying to figure out impact of off/on campus

```{r}
data |>
  filter(training_holdout == "training") |>
  distinct(student_id, .keep_all = TRUE) |>
  count(off_campus, distance_zero = distance_to_campus == 0)

```

# graphing all 4 coverages for distance to campus
```{r}
distance_all_coverages <- data |>
  filter(training_holdout == "training") |>
  mutate(distance_group = cut(
    distance_to_campus,
    breaks = c(-Inf, 0, 1, 3, 6, Inf),
    labels = c("0", "0-1", "1-3", "3-6", "6+")
  )) |>
  group_by(student_id, coverage, risk_tier, distance_group) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, distance_group) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

ggplot(
  distance_all_coverages,
  aes(
    x = distance_group,
    y = avg_amount_per_policyholder
  )
) +
  geom_col() +
  facet_grid(coverage ~ risk_tier, scales = "free_y") +
  labs(
    title = "Average Loss per Policyholder by Distance, Coverage, and Risk Tier",
    x = "Distance to Campus",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal()

distance_all_coverages |>
  select(
    coverage,
    risk_tier,
    distance_group,
    num_policyholders,
    num_with_claim
  )


```

# SPRINKLERED

```{r}

sprinkler_summary <- data |>
  filter(training_holdout == "training") |>
  group_by(student_id, coverage, risk_tier, sprinklered) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, sprinklered) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

sprinkler_plot <- ggplot(
  sprinkler_summary,
  aes(x = sprinklered, y = avg_amount_per_policyholder)
) +
  geom_col() +
  facet_grid(
    coverage ~ risk_tier,
    scales = "free_y",
    labeller = labeller(
      coverage = c(
        "Additional Living Expense" = "ALE",
        "Guest Medical" = "GM",
        "Liability" = "Liability",
        "Personal Property" = "PP"
      )
    )
  ) +
  labs(
    title = "Average Loss per Policyholder by Sprinkler Status, Coverage, and Risk Tier",
    x = "Sprinklered",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal()

sprinkler_plot

sprinkler_summary |>
  select(
    coverage,
    risk_tier,
    sprinklered,
    num_policyholders,
    num_with_claim
  )

ggsave(
  "~/Desktop/sprinkler_status.png",
  plot = sprinkler_plot,
  width = 12,
  height = 7,
  dpi = 300,
  bg = "white"
)

```
# GREEK LIFE

```{r}
greek_summary <- data |>
  filter(training_holdout == "training") |>
  group_by(student_id, coverage, risk_tier, greek) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, greek) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

ggplot(
  greek_summary,
  aes(x = greek, y = avg_amount_per_policyholder)
) +
  geom_col() +
  facet_grid(
    coverage ~ risk_tier,
    scales = "free_y",
    labeller = labeller(
      coverage = c(
        "Additional Living Expense" = "ALE",
        "Guest Medical" = "GM",
        "Liability" = "Liability",
        "Personal Property" = "PP"
      )
    )
  ) +
  labs(
    title = "Average Loss per Policyholder by Greek Status, Coverage, and Risk Tier",
    x = "Greek Status",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal()

greek_summary |>
  select(
    coverage,
    risk_tier,
    greek,
    num_policyholders,
    num_with_claim
  )

ggsave(
  "greek_status.png",
  width = 12,
  height = 7,
  dpi = 300,
  bg = "white"
)

getwd()
```

# CLASS

```{r}

class_summary <- data |>
  filter(training_holdout == "training") |>
  group_by(student_id, coverage, risk_tier, class) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, class) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

ggplot(
  class_summary,
  aes(
    x = class,
    y = avg_amount_per_policyholder
  )
) +
  geom_col() +
  facet_grid(coverage ~ risk_tier, scales = "free_y") +
  labs(
    title = "Average Loss per Policyholder by Class, Coverage, and Risk Tier",
    x = "Class",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

class_summary |>
  select(
    coverage,
    risk_tier,
    class,
    num_policyholders,
    num_with_claim
  )

```
```{r}
table(data$class, data$risk_tier)
```
# investigating off campus and distance to campus

```{r}
data |>
  filter(training_holdout == "training") |>
  distinct(student_id, off_campus, distance_to_campus) |>
  group_by(off_campus) |>
  summarize(
    num_students = n(),
    min_distance = min(distance_to_campus),
    max_distance = max(distance_to_campus),
    mean_distance = mean(distance_to_campus)
  )

offcampus_summary <- data |>
  filter(training_holdout == "training") |>
  group_by(student_id, coverage, risk_tier, off_campus) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, off_campus) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

offcampus_summary

offcampus_plot <- ggplot(
  offcampus_summary,
  aes(x = off_campus, y = avg_amount_per_policyholder)
) +
  geom_col() +
  facet_grid(
    coverage ~ risk_tier,
    scales = "free_y",
    labeller = labeller(
      coverage = c(
        "Additional Living Expense" = "ALE",
        "Guest Medical" = "GM",
        "Liability" = "Liability",
        "Personal Property" = "PP"
      )
    )
  ) +
  labs(
    title = "Average Loss per Policyholder by Campus Status, Coverage, and Risk Tier",
    x = "Campus Status",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal()

offcampus_plot

ggsave(
  "~/Desktop/campus_status.png",
  plot = offcampus_plot,
  width = 12,
  height = 7,
  dpi = 300,
  bg = "white"
)


```
# STUDY (aka MAJOR)

```{r}
data |>
  filter(training_holdout == "training") |>
  distinct(student_id, study) |>
  count(study, sort = TRUE)

study_summary <- data |>
  filter(training_holdout == "training") |>
  group_by(student_id, coverage, risk_tier, study) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, study) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = mean(total_amount[had_claim]),
    .groups = "drop"
  )

ggplot(
  study_summary,
  aes(
    x = study,
    y = avg_amount_per_policyholder
  )
) +
  geom_col() +
  facet_grid(coverage ~ risk_tier, scales = "free_y") +
  labs(
    title = "Average Loss per Policyholder by Major, Coverage, and Risk Tier",
    x = "Major",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

study_summary |>
  select(
    coverage,
    risk_tier,
    study,
    num_policyholders,
    num_with_claim
  )
```
# GENDER

```{r}
data |>
  distinct(student_id, gender) |>
  count(gender, sort = TRUE)

gender_summary <- data |>
  group_by(student_id, coverage, risk_tier, gender) |>
  summarize(
    total_amount = sum(amount),
    had_claim = any(amount > 0),
    .groups = "drop"
  ) |>
  group_by(coverage, risk_tier, gender) |>
  summarize(
    num_policyholders = n(),
    num_with_claim = sum(had_claim),
    claim_frequency = mean(had_claim),
    avg_amount_per_policyholder = mean(total_amount),
    avg_loss_per_claimant = ifelse(
      num_with_claim > 0,
      mean(total_amount[had_claim]),
      NA
    ),
    .groups = "drop"
  )

gender_summary

ggplot(
  gender_summary,
  aes(x = gender, y = avg_amount_per_policyholder)
) +
  geom_col() +
  facet_grid(
    coverage ~ risk_tier,
    scales = "free_y"
  ) +
  labs(
    title = "Average Loss per Policyholder by Gender, Coverage, and Risk Tier",
    x = "Gender",
    y = "Average Loss per Policyholder"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
















