---
title: "Pure Premium Modeling"
date: "2026-09-14"
output:
  html_document:
    toc: true
    toc_float: true
---

Source: https://metricgate.com/blogs/tweedie-glm-insurance-pure-premium/

# Setting Up

```{r}
# load packages
library(tidyverse)
library(statmod)
library(tweedie)

# load dataset
data <- read_csv("06 - CAS Predictive Modeling Case Competition- Dataset.csv")
head(data)
```

# Data Preprocessing

The given data set has 40,071 rows, but we are told that there are only 10,000 policyholders and 4 different coverage types. Since there's no duplicated rows, this means there are some policyholders who filed more than one claim. Thus, for policyholders who filed more than one claim per insurance coverage, we need to combine the amount for those claims. 

```{r}
# check if there's any duplicate rows
cat("There are ", sum(duplicated(data)), "duplicated rows.")
```

```{r}
# converting risk tier into categorical variable
data$risk_tier <- as.character(data$risk_tier)
head(data)
```

```{r}
#Combining Multiple Claims made by the same policyholder and coverage
data <- data |>
  group_by(student_id,coverage) |>
  summarize(
    amount= sum(amount, na.rm = TRUE),
    name= first(name),
    class=first(class),
    study=first(study),
    gpa=first(gpa),
    greek= first(greek),
    off_campus= first(off_campus),
    distance_to_campus = first(distance_to_campus),
    gender= first(gender),
    sprinklered =first(sprinklered),
    risk_tier= first(risk_tier),
    training_holdout= first(training_holdout),
    .groups = "drop"
)

head(data)
```

```{r}
# Grouping the continous variables into 3 bands to make the pricing model easier to rate 
data <- data |>
  mutate(
    gpa_interval= case_when(
      gpa<1 ~ "Sub 1.0",
      gpa<3 ~ "1.0 to 2.99",
      gpa>=3 ~"Greater than 3"
    ),
    
    distance_interval= case_when(
    distance_to_campus==0 ~ "Living On Campus",
    distance_to_campus <=1 ~ "Living 1 mile or less from Campus",
    distance_to_campus > 1 ~ " Living more than 1 mile from Campus"
  )
)

```

# 1. Personal Property

## Full Model

```{r}
training <- data |> filter(training_holdout == "training", 
                           coverage == "Personal Property")
holdout <- data |> filter(training_holdout == "holdout",
                          coverage == "Personal Property")
```

```{r}
# estimate Tweedie power index p

profile <- tweedie_profile(
  amount ~ risk_tier + study + gpa + greek + off_campus +
    distance_to_campus + gender + sprinklered,
  data = training, p.vec = seq(1.1, 1.6, by = 0.1),
  link.power = 0, do.plot = FALSE
)

p_hat <- profile$p.max
cat("Estimated Tweedie power p:", round(p_hat, 3))

# fits Tweedie GLM w/ log link
personal_prop_full_model <- glm(
  amount ~ risk_tier + study + gpa + greek + off_campus + 
    distance_to_campus + gender + sprinklered,
  data = training,
  family = tweedie(var.power = p_hat, link.power = 0),
  control = glm.control(maxit = 15)
)

summary(personal_prop_full_model)
```

```{r}
holdout$pred  <- predict(personal_prop_full_model, newdata = holdout, type = "response")

personal_prop_full_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
personal_prop_full_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Average Personal Property Amount: \n Actual vs. Predicted \n (Full Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 120),
        ylab = "Average Value")
```

```{r}
personal_prop_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
personal_prop_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
personal_prop_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Personal Property Claim Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

## Variable Selection

Since there are more samples than variables ($n = 40071 > p = 11$), we use Backward Stepwise Selection to perform variable selection. We choose a p-value of 0.05 as our cutoff. 

### Iteration 1

Based on the table below, we should drop the variable `distance_to_campus`. 

```{r}
# removing 1 variable
drop1(personal_prop_full_model, test = "F")
```

```{r}
personal_prop_model_1 <- update(personal_prop_full_model, . ~ . - distance_to_campus)
summary(personal_prop_model_1)
```

### Iteration 2

Based on the table below, we should drop the variable `gender` next. 

```{r}
drop1(personal_prop_model_1, test = "F")
```

```{r}
personal_prop_model_2 <- update(personal_prop_model_1, . ~ . - gender)
summary(personal_prop_model_2)
```

### Iteration 3

Based on the table below, we should drop the variable `gpa` next. 

```{r}
drop1(personal_prop_model_2, test = "F")
```

```{r}
personal_prop_model_3 <- update(personal_prop_model_2, . ~ . - gpa)
summary(personal_prop_model_3)
```

### Iteration 4

Using a p-value cutoff of 0.05, the table below suggests we should stop removing variables. Thus, `personal_prop_model_3` is our final model for predicting the expected claim loss for personal property claims. 

```{r}
drop1(personal_prop_model_3, test = "F")
```

## Final Model

```{r}
holdout$pred  <- predict(personal_prop_model_3, newdata = holdout, type = "response")

personal_prop_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
personal_prop_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Average Personal Property Amount: \n Actual vs. Predicted \n (Reduced Model)",
        names.arg = c("Actual", "Predicted"), 
        ylab = "Average Value",
        ylim = c(0, 120))
```

```{r}
personal_prop_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
personal_prop_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
personal_prop_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Personal Property Claim Amount (Reduced Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

# 2. Additional Living Expenses

## Full Model

```{r}
training <- data |> filter(training_holdout == "training", 
                           coverage == "Additional Living Expense")
holdout <- data |> filter(training_holdout == "holdout",
                          coverage == "Additional Living Expense")
```

```{r}
# estimate Tweedie power index p

profile <- tweedie_profile(
  amount ~ risk_tier + study + gpa + greek + off_campus +
    distance_to_campus + gender + sprinklered,
  data = training, p.vec = seq(1.1, 1.6, by = 0.1),
  link.power = 0, do.plot = FALSE
)

p_hat <- profile$p.max
cat("Estimated Tweedie power p:", round(p_hat, 3))

# fits Tweedie GLM w/ log link
living_exp_full_model <- glm(
  amount ~ risk_tier + study + gpa + greek + off_campus + 
    distance_to_campus + sprinklered + risk_tier,
  data = training,
  family = tweedie(var.power = p_hat, link.power = 0),
  control = glm.control(maxit = 15)
)
```

```{r}
holdout$pred  <- predict(living_exp_full_model, newdata = holdout, type = "response")

living_exp_full_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
living_exp_full_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Average Additional Living Expenses: \n Actual vs. Predicted \n (Full Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 325),
        ylab = "Average Value")
```

```{r}
living_exp_full_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
living_exp_full_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
living_exp_full_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Additional Living Expenses Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

## Variable Selection

### Iteration 1

Based on the table below, we should first drop the variable `distance_to_campus`. 

```{r}
drop1(living_exp_full_model, test = "F")
```

```{r}
living_exp_model1 <- update(living_exp_full_model, ~ . - distance_to_campus)
summary(living_exp_model1)
```

### Iteration 2

Based on the table below, we should drop the variable `study` next. 

```{r}
drop1(living_exp_model1, test = "F")
```

```{r}
living_exp_model2 <- update(living_exp_model1, ~ . - study)
summary(living_exp_model2)
```

### Iteration 3

Using a p-value cutoff of 0.05, the table below suggests we should stop removing variables.

```{r}
drop1(living_exp_model2, test = "F")
```

## Final Model

```{r}
holdout$pred  <- predict(living_exp_model2, newdata = holdout, type = "response")

living_exp_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
living_exp_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Average Additional Living Expenses Amount: \n Actual vs. Predicted \n (Reduced Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 325),
        ylab = "Average Value")
```

```{r}
living_exp_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
living_exp_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
living_exp_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Additional Living Expenses Amount (Reduced Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

# 3. Liability Protection

## Full Model

```{r}
training <- data |> filter(training_holdout == "training", 
                           coverage == "Liability")
holdout <- data |> filter(training_holdout == "holdout",
                          coverage == "Liability")
```

```{r}
# estimate Tweedie power index p

profile <- tweedie_profile(
  amount ~ risk_tier + study + gpa + greek + off_campus +
    distance_to_campus + gender + sprinklered,
  data = training, p.vec = seq(1.1, 1.6, by = 0.1),
  link.power = 0, do.plot = FALSE
)

p_hat <- profile$p.max
cat("Estimated Tweedie power p:", round(p_hat, 3))

# fits Tweedie GLM w/ log link
liability_full_model <- glm(
  amount ~ risk_tier + study + gpa + greek + off_campus + 
    distance_to_campus + gender + sprinklered,
  data = training,
  family = tweedie(var.power = p_hat, link.power = 0),
  control = glm.control(maxit = 15)
)

summary(liability_full_model)
```

```{r}
holdout$pred  <- predict(liability_full_model, newdata = holdout, type = "response")

liability_full_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
liability_full_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Liability Amount: \n Actual vs. Predicted \n (Full Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 150),
        ylab = "Average Value")
```

```{r}
liability_full_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
liability_full_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
liability_full_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Liability Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

## Variable Selection

### Iteration 1

The table below suggests we drop the variable `gpa`. 

```{r}
drop1(liability_full_model, test = "F")
```

```{r}
liability_model1 <- update(liability_full_model, ~ . - gpa)
summary(liability_model1)
```

### Iteration 2

The table below suggests we stop dropping variables. So `liability_mdoel1` is our final model. 

```{r}
drop1(liability_model1, test = "F")
```

## Final Model

```{r}
holdout$pred  <- predict(liability_model1, newdata = holdout, type = "response")

liability_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
liability_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Liability Amount: \n Actual vs. Predicted \n (Reduced Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 150),
        ylab = "Average Value")
```

```{r}
liability_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
liability_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
liability_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Liability Claim Amount (Reduced Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

# 4. Guest Medical Protection

## Full Model

```{r}
training <- data |> filter(training_holdout == "training", 
                           coverage == "Guest Medical")
holdout <- data |> filter(training_holdout == "holdout",
                          coverage == "Guest Medical")
```

```{r}
# estimate Tweedie power index p

profile <- tweedie_profile(
  amount ~ risk_tier + study + gpa + greek + off_campus +
    distance_to_campus + gender + sprinklered,
  data = training, p.vec = seq(1.1, 1.6, by = 0.1),
  link.power = 0, do.plot = FALSE
)

p_hat <- profile$p.max
cat("Estimated Tweedie power p:", round(p_hat, 3))

# fits Tweedie GLM w/ log link
guest_medical_full_model <- glm(
  amount ~ risk_tier + study + gpa + greek + off_campus + 
    distance_to_campus + gender + sprinklered,
  data = training,
  family = tweedie(var.power = p_hat, link.power = 0),
  control = glm.control(maxit = 15)
)

summary(guest_medical_full_model)
```

```{r}
holdout$pred  <- predict(guest_medical_full_model, newdata = holdout, type = "response")

guest_medical_full_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
guest_medical_full_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Guest Medical Amount: \n Actual vs. Predicted \n (Full Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 500),
        ylab = "Average Value")
```

```{r}
guest_medical_full_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
guest_medical_full_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
guest_medical_full_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Guest Medical Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```

## Variable Selection

### Iteration 1

The table below suggests we drop the variable `gender`. 

```{r}
drop1(guest_medical_full_model, test = "F")
```

```{r}
guest_medical_model1 <- update(guest_medical_full_model, ~ . - gender)
summary(guest_medical_model1)
```

### Iteration 2

The table below suggests we stop dropping variables. 

```{r}
drop1(guest_medical_model1, test = "F")
```

## Final Model 

```{r}
holdout$pred  <- predict(guest_medical_model1, newdata = holdout, type = "response")

guest_medical_avg_amt <- holdout |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
guest_medical_avg_amt
```

```{r}
col_means <- colMeans(holdout[c("amount", "pred")])

barplot(col_means, 
        col = c("goldenrod1", "royalblue4"),
        main = "Guest Medical Amount: \n Actual vs. Predicted \n (Reduced Model)",
        names.arg = c("Actual", "Predicted"), 
        ylim = c(0, 500),
        ylab = "Average Value")
```

```{r}
guest_medical_avg_amt <- holdout |>
  group_by(risk_tier) |>
  summarize(observed = mean(amount), 
            fitted = mean(pred),
            abs_difference = abs(observed - fitted))
guest_medical_avg_amt
```

```{r}
# barplot to compare actual average claim amount vs average predicted claim amount
guest_medical_avg_amt |>
  pivot_longer(cols = c(observed, fitted),
               names_to = "type", values_to = "mean_pp") |>
  mutate(type = factor(type, levels = c("observed", "fitted"),
                       labels = c("Observed", "Fitted"))) |>
  ggplot(aes(x = risk_tier, y = mean_pp, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Observed" = "royalblue4",
                               "Fitted" = "goldenrod1")) +
  labs(title = "Observed vs. Tweedie-fitted \n Average Guest Medical Amount (Full Model)",
       x = "Risk Tier", y = "Mean Pure Premium", fill = "")
```



